```{r setup, include = FALSE}
library(scales)
library(tidyverse)
knitr::opts_chunk$set(
  caching = TRUE,
  collapse = TRUE
)
```

### Loading the data

```{r, echo=FALSE}
data_frame = read_csv('./gapminder_clean.csv', na = c('', 'N/A'))
```


```{r}
head(data_frame)
```

### Data Visualization
```{r}

filtered_data = data_frame |>
  filter(Year == 1962) 
  

filtered_data |>
  select(starts_with('CO2'), gdpPercap) |>
  rename(co2 = starts_with('CO2')) |>
  filter(if_all(everything(), ~ !is.na(.))) |>
  ggplot() +
  geom_point(aes(x = gdpPercap, y = co2)) +
  labs(
    x = 'GDP per capita',
    y = 'CO2 emissions (metric tons per capita)',
    title = 'CO2 emissions per capita generally increases with GDP'
  )
```



### correlation between CO2 emission and GDP growth
```{r}
correlation_columns = filtered_data |>
  select(starts_with('CO2'), gdpPercap) |>
  rename(co2 = starts_with('CO2')) |>
  filter(if_all(everything(), ~ !is.na(.)))

#since the tow variables are quantitative continuous, i used Pearson correlation
#by default the cor() function will use Pearson method
#the function output a value of 0.9260817 indicating there is a positive relation between the two variables (evolving in the same direction)

cor(correlation_columns$co2,
    correlation_columns$gdpPercap,
    method = 'pearson')

#the p_vlaue is used to assess the probability of getting a correlation coefficient as extreme as r, if we sample from a population when the null hypothesis r = 0 is true

cor.test(correlation_columns$co2,
    correlation_columns$gdpPercap,
    method = 'pearson')
```

```{r}

data_frame |>
  select(
    Year,
    co2 = starts_with('CO2'),
    gdpPercap
  ) |>
  filter(!is.na(co2), !is.na(gdpPercap)) |>
  group_by(Year) |>
  summarise(
    r = cor(co2, gdpPercap),
    .groups = 'drop'
  ) |>
  ggplot() +
  geom_bar(aes(x = Year, y = r),
           stat = 'identity') +
  geom_text(aes(x = Year, y = r, label = round(r, 2)),
                vjust = -0.5) +
  scale_x_continuous(breaks = sort(unique(data_frame$Year))) +
  theme(legend.position = 'none')
```

```{r}
library(plotly)
plot = data_frame |>
  filter(Year == 1967) |>
  select( co2 = starts_with('CO2'),
          gdpPercap,
          pop,
          continent) |>
  ggplot() +
  geom_point(aes(x = gdpPercap, y = co2, size = pop, color = continent)) +
  scale_x_continuous(
    labels = label_dollar(scale = 1/1000, suffix = 'K')
  ) +
  coord_cartesian(
    xlim = c(0,20000),
    ylim = c(0,25)
  ) +
  labs(
    x = 'GDP per capita',
    y = 'CO2 emissions (metric tons per capita)'
    )

ggplotly(plot) 
```
### stat test 

##What is the relationship between continent and 'Energy use (kg of oil equivalent per capita)'? (stats test needed)


```{r}
### box plot for energy use accross continent 
### run a linear regression model 
### run anova

fit = lm(`Energy use (kg of oil equivalent per capita)`~continent, data = data_frame)
anova(fit)
```

I used ANOVA to test whether there are significant differences in energy use across continents, under the assumption that the average energy use is equal for all continents. 

Null Hypothesis H0: there isn't a difference in energy use across continent

Alternative Hypothesis H1: at least one group mean energy use is different 

the threshold for p-value is set to 0.05

The fourth column displays the p-value, which indicates the probability—under the null hypothesis—of observing an F-value as extreme as the one obtained. the p_value corresponds to 8.53 × 10⁻³, thus we reject the null hypothesis.

if we want to compare individual groups, we can use a contrast matrix.

##Is there a significant difference between Europe and Asia with respect to 'Imports of goods and services (% of GDP)' in the years after 1990? (stats test needed)

since we are comparing two independents groups. we will use t-test for comparsion of group means.

Null Hypothesis: There is no significant difference in imports as a percentage of GDP between Asia and Europe.
Alternative Hypothesis: The mean import percentage of GDP differs between Asia and Europe.


```{r}

euro_asia = data_frame |>
  filter(Year > 1990) |>
  filter(continent == 'Asia' | continent == 'Europe') |>
  rename(import_goods = starts_with('Import')) |>
  select(continent, import_goods) |>
  filter(if_all(everything(), ~ !is.na(import_goods)))


t_test = t.test(import_goods ~ continent, data = euro_asia)$p.value

```

since (p > 0.05), we will accept the Null Hypothesis that there isn't a significant difference between Asia and Europe in terms of Import of Goods and Services as a percentage of GDP.

##What is the country (or countries) that has the highest 'Population density (people per sq. km of land area)' across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)

```{r}

data_frame |>
  group_by(`Country Name`) |>
  summarise(avg_pop_dens_yrs = as.numeric(mean(`Population density (people per sq. km of land area)`, na.rm = TRUE))) |>
  arrange(desc(avg_pop_dens_yrs)) |>
  head(10) |>
  ggplot() +
  geom_bar(aes(x = `Country Name`, y = avg_pop_dens_yrs),
           stat = 'identity') +
  geom_text(aes(x = `Country Name`, y = avg_pop_dens_yrs, 
                 label = round(avg_pop_dens_yrs, 2)),
            vjust = -0.25) +
  coord_cartesian(ylim = c(0,17000))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
as we can observe in the bar chart, Macao region in China has the highest average population density over the years.

## What country (or countries) has shown the greatest increase in 'Life expectancy at birth, total (years)' between 1962 and 2007?

```{r}

data_frame |>
  group_by(`Country Name`) |>
  filter(Year %in% c(1962:2007)) |>
  summarise(
    birth_exp_2007 = `Life expectancy at birth, total (years)`[Year == 2007],
    birth_exp_1962 = `Life expectancy at birth, total (years)`[Year == 1962],
    percent_change = ((birth_exp_2007 - birth_exp_1962) / birth_exp_1962) * 100
                         ) |>
  arrange(desc(percent_change)) |>
  head(10) |>
  ggplot() +
  geom_bar(aes(x = `Country Name`, y = percent_change),
           stat = 'identity') +
  geom_text(aes(x = `Country Name`, y = percent_change, 
                 label = round(percent_change, 2)),
            vjust = -0.25) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    y = 'Aver'
  )
```
